---
title: "Introduction to R"
subtitle: "Homework 3"
author: "Koech Andrew"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
mainfont: Calibri Light
output: pdf_document
geometry: margin=1in
linkcolor: blue
urlcolor: blue

header-includes:
  - \usepackage{amsfonts}
  - \usepackage{amssymb}
  - \usepackage{array}
  - \usepackage{booktabs}
  - \usepackage{mathtools}
  - \usepackage{braket}
---
\newpage

```{r,echo=FALSE,include=FALSE}
# options(digits = 4)
```

## Homework task 1

Let the height (in cm) of a population be normally distributed with $\mu = 170$ and $\sigma = 10$. Set the seed 1528 and use the following command to generate a sample of size n = 2000 from this population: pop1<-rnorm(2000,170,10).


a) Round the height to integers by using round() (and overwrite the values in pop1).

```{r}
set.seed(1528)
pop1<-rnorm(2000,170,10)
head(pop1)
pop1<-round(pop1,digits = 0)
head(pop1)
```

b) Execute which(pop1 > 190) and interpret the result.

```{r}
which(pop1 > 190)# The output here are the indices of the values that meet the criteria that is heights above 190cm 
```

c) What subgroup is created by running pop1[pop1 > 190]?

```{r}
pop1[pop1 > 190]#A subgroup of heights above 190cm ONLY are selected.
```

d) Create two subgroups: 1) height < 155 and 2) height > 185.

```{r}
height_less155<-pop1[pop1 < 155]
height_greater185<-pop1[pop1 > 185]
```


e) Interpret the results:

```{r,warning=FALSE}
rev(pop1)# This function reverses the order of height values from the last to first

unique(pop1)# This function gives a list of unique height values from pop1 vector since we have a repetition of this values more than once in the vector meaning we have 60 unique values in the vector pop1

duplicated(pop1)#This function checks for duplicated height values and returns a logical output such that the decision returned is by comparing the second value to the first where if the value is unique then the result is FALSE otherwise it would be TRUE.

pop1[duplicated(pop1)]#This function creates a subgroup of the height values that are duplicates of pop1

pop1[!duplicated(pop1)]#This function creates a subgroup of the height values that are unique values of pop1
```

f) Simulate another sample pop2 of size n = 1000 from a population whose height is N($\mu = 172, \sigma = 12$)-distributed.

```{r}
set.seed(1528)
pop2<-rnorm(1000,172,12)
```

g) Round the height in the second population to integers.

```{r}
head(pop2)
pop2<-round(pop2,digits = 0)
head(pop2)
```

h) **Randomly** assign 1000 individuals in sample 1 to those in sample 2 by defining a matrix X with two columns, one containing the individuals from sample 1 and one containing the individuals from sample 2 that the individuals from sample 1 were assigned to. _Hint: Use sample() for the randomization and cbind() for the assignment._

```{r}
X<-cbind(pop1=sample(pop1, 1000, replace = FALSE),pop2)
```

i) We consider each row of the matrix X as a pair, where individuals from sample 1 and 2 are seen as height values by men and women, respectively. How many pairs are there where the man is shorter than the woman?

```{r}
length(which(X[,1]<X[,2]))# 533 pairs meet the criteria where the man is shorter than the woman
```

## Homework task 2 

a) Write a function getOdd that returns all the odd numbers in a numeric-vector x. This function should first check whether the vector is indeed a vector and of type numeric. If x does not meet these two requirements return the string "x has to be a numeric vector!", otherwise convert the numeric vector to an integer and return a vector of all odd numbers that occur in x without duplicates (see unique() from task 1). Note: An integer (a whole number) $z \in Z$ is odd if z modulo 2 = 1 (so if z/2 has a remainder = 1).

```{r}
getOdd<-function(x){
  if(w<-(is.vector(x) & is.numeric(x))==FALSE){
    return("x has to be a numeric vector!")
  }else{
    w<-as.integer(x)
  }
  y<-which(w%%2==1)
  z<-w[y]
  return(unique(z))
}
```


b) Generate a random sample of 50 numbers from the set ${-50, -49,..., 0,..., 49, 50}$ using sample() (without replacement). Check whether your sample contains any negative values. Also, apply the function from a) to your sample and from the resulting vector extract all numbers that are divisible by 3 (all z with z mod 3 = 0).

```{r}

k<-sample(-50:50,50,replace = FALSE)
k#random sample of 50 numbers from {-50,...,49,50}

odd<-getOdd(k)# applying the getOdd function

#extracting all numbers that are divisible by 3
odd[which(odd%%3==0)]
```

## Homework task 3

The variable Wind in airquality gives the wind speed in mph (mile per hour). This is a metric-scaled variable. Form a new variable (ordinal), Wind, from this variable: and add this variable to the airquality data set. Now define the ordinal-scaled feature Wind3 like this:

```{r}
data("airquality")
airquality$Wind2 <- ifelse(airquality$Wind <= 12, "weak", "strong")

airquality$Wind3 <- ifelse(airquality$Wind <= 4, "weak",
                           ifelse(airquality$Wind <= 18, "mild", "strong"))
```

